---
title: "Reading in Data and Calculating Prevalence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading in Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(STAVE)
if (requireNamespace("kableExtra", quietly = TRUE)) {
    library(kableExtra)
}
```

This tutorial covers the following topics:

- Creating a new STAVE object
- Loading data into this object
- Calculating prevalence, with and without ambiguous matches

## Reading in data

Our input data will take the form of three tables. We will use the example data that comes loaded with the STAVE package. These tables are already correctly formatted. If you are using your own data then you need to match this input format (see [explanation](articles/input_formats.html)):

```{r}
data("example_input")

example_input$studies |>
  kbl(format = "html", table.attr = "style='width:100%; white-space: nowrap;'") |>
  scroll_box(width = "100%", height = NULL)

example_input$surveys |>
  kbl(format = "html", table.attr = "style='width:100%; white-space: nowrap;'") |>
  scroll_box(width = "100%", height = NULL)

example_input$counts |>
  kbl(format = "html", table.attr = "style='width:100%; white-space: nowrap;'") |>
  scroll_box(width = "100%", height = NULL)
```

With the data correctly formatted, we can create a STAVE object and append the data:
```{r}
# create new object
s <- STAVE_object$new()

# append data
s$append_data(studies_dataframe = example_input$studies,
              surveys_dataframe = example_input$surveys,
              counts_dataframe = example_input$counts)

# check how many studies are now loaded
s
```

Once data are loaded, we can always view the different tables using *get* functions. However, we cannot alter the values directly.

As a side note, if we want to know all the variants in our loaded data, we can use the `get_variants()` function.

```{r}
s$get_variants()
```

## Calculating prevalence

We can calculate the prevalence of any variant using `get_prevalence()`. This appends all information together over studies and surveys, with final columns giving the prevalence (%) and lower and upper 95% CIs:
```{r}
s$get_prevalence("k13:469:Y") |>
  kbl(format = "html", table.attr = "style='width:100%; white-space: nowrap;'") |>
  scroll_box(width = "100%", height = NULL)
```

Notice that results are given for every loaded survey, even if there is no corresponding genetic data to calculate a prevalence. If instead you only want rows for which there is a non-zero denominator, use the argument `return_full = FALSE`.

Here is another example, this time allowing for ambiguous matches. 
```{r}
s$get_prevalence("k13:469:Y", keep_ambiguous = TRUE, prev_from_min = TRUE) |>
  kbl(format = "html", table.attr = "style='width:100%; white-space: nowrap;'") |>
  scroll_box(width = "100%", height = NULL)
```

A `min` and `max` numerator are now given. There is still only a single prevalence estimate and 95% CI, calculated using either the `min` or the `max` values, specified by the `prev_from_min` argument.

