---
title: "Reading in Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading in Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, echo=FALSE}
library(STAVE)
if (requireNamespace("kableExtra", quietly = TRUE)) {
    library(kableExtra)
}
```

This tutorial covers the following topics:

- Creating a new STAVE object
- Loading data into this object
- Viewing the loaded data
- Dropping studies and surveys by ID

## Reading in data

STAVE works via a single class (an R6 object) that acts as the main data container.
This class allows users to efficiently import, store, and manipulate
data via specialized member functions.

A new object can be created and data read in like this:

```{r}
# create new object
s <- STAVE_object$new()

# append data using a member function
s$append_data(studies_dataframe = example_input$studies,
              surveys_dataframe = example_input$surveys,
              counts_dataframe = example_input$counts)
```

The default print method tells us how many studies and surveys are loaded:

```{r}
s
```

This example uses the `example_input` object that comes pre-loaded with the
package - have a look at this object to get a feel for the required input format (and see the rest of this package documentation for a rigorous specification).

Using a custom class offers several key advantages. Once loaded, all data remain
consolidated within a single object, avoiding fragmentation. The class structure
also ensures the data are *encapsulated*, meaning they cannot be directly edited
by the user. This built-in protection minimizes the risk of accidental data
corruption.

We can view the loaded tables using `get` functions:

```{r, eval=FALSE}
s$get_studies()
```

```{r, echo=FALSE}
s$get_studies() |>
  kbl(format = "html", table.attr = "style='width:100%; white-space: nowrap;'") |>
  scroll_box(width = "100%", height = NULL)
```

```{r, eval=FALSE}
s$get_surveys()
```

```{r, echo=FALSE}
s$get_surveys() |>
  kbl(format = "html", table.attr = "style='width:100%; white-space: nowrap;'") |>
  scroll_box(width = "100%", height = NULL)
```

```{r, eval=FALSE}
s$get_counts()
```

```{r, echo=FALSE}
s$get_counts() |>
  kbl(format = "html", table.attr = "style='width:100%; white-space: nowrap;'") |>
  scroll_box(width = "100%", height = NULL)
```

However, we cannot directly modify the data in these tables. Instead, we have to drop studies and surveys using member functions.

## Dropping studies and surveys

Imagine we are not interested in the study by Dama et al. (2017), and want to drop it from our analysis. We can do so via the `study_id`:

```{r}
s$drop_study("Dama_2017")
```

Similarly, we may want to drop a specific survey - for example the `Asua_2019_Agago_2017` survey:

```{r}
s$drop_survey("Asua_2019_Agago_2017")
```

Looking at the STAVE object, we can see that we have reduced the size of the data:

```{r}
s
```


