---
title: "tutorial_basic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tutorial_basic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(STAVE)
```

## Example workflow

Input data take the form of three tables. We will use the pre-loaded example
data here, if you are using your own data then you need to match this input format:
```{r}
library(STAVE)

data("example_input")
#View(example_input$studies)
#View(example_input$surveys)
#View(example_input$counts)
```

Create a STAVE object and append the data:
```{r}
# create new object
s <- STAVE_object$new()

# append data using a member function
s$append_data(studies_dataframe = example_input$studies,
              surveys_dataframe = example_input$surveys,
              counts_dataframe = example_input$counts)

# check how many studies are now loaded
s
```

Once data are loaded, we can always view the different tables using *get* functions. However, we cannot alter the values directly.
```{r}
s$get_studies() |> head()
s$get_surveys() |> head()
s$get_counts() |> head()
```

We can calculate the prevalence of any variant using `get_prevalence()`:
```{r}
p <- s$get_prevalence("mdr1:184:F")
#View(p)
```

We can also return a list of all variants in the data:
```{r}
s$get_variants()
```

Finally, we can selectively drop studies from the data using their study ID:
```{r}
s$drop_study(drop_study_ID = "wwarn_10297_Nelson")
s
```
